---
title: "Chapter 4"
output:
  html_notebook: 
    highlight: tango
    theme: spacelab
    code_folding: show
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, include = TRUE)
```


```{r, include=FALSE}
library(tidyverse)
```

# General Notes
- Keyboard shortcut for assignment operator: Alt + -
- R is case sensitive
- You can immediately print assignment results to screen by surrounding assignment with parentheses

```{r}
(y <- seq(1, 10))
```

## Exercises
1. Why does this code not work?

    The example has an alternative character in place of the 'i' in the assignment statement.

2. Tweak each of the following R commands so they run correctly.
```{r}
#ggplot(dota = mpg) + 
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))



#fliter(mpg, cyl = 8)
filter(mpg, cyl == 8)

#filter(diamond, carat > 3)
filter(diamonds, carat > 3)
```


3. Press Alt + Shift + K. What happens? How can you get to the same place using the menus?

    This keyboard shortcut toggles a pop-up showing the keyboard shortcut quick reference. You can also access the available keyboard shortcuts by using the menu options 'Tools' > 'Keyboard Shortcuts Help'.