---
title: "Chapter 3"
output:
  html_notebook: 
    highlight: tango
    theme: spacelab
    code_folding: show
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, include = TRUE)
```


```{r, include=FALSE}
library(tidyverse)
```

# 3.2 First Steps
```{r view}
head(mpg) # preview data
```


```{r plot}
ggplot(data = mpg) +
    geom_point(mapping = aes(x = displ, y = hwy))
```


##Exercises
1. What happens with the following code ```ggplot(data = mpg)```?

    Nothing - the ggplot() function just creates the coordinate system for later plots.

2. How many rows are in mpg? How many columns?
```{r}
nrow(mpg)
```
```{r}
ncol(mpg)
```

3. What does the ```drv``` variable describe?
      
      ```drv``` describes whether or not a car has front, rear, or four wheel drive.

```{r}
?mpg
```


4. Make a scatterplot of hwy vs cyl.
```{r}
ggplot(data = mpg) +
  geom_point(aes(x = cyl, y = hwy))
```

5. What happens if you make a scatterplot of ```class``` and ```drv```? Why is the plot not useful?

    The plot isn't useful as a scatterplot, because it is mapping two categorical variables against one another on a plot that is intended to represent quantitative scale -- the differences between steps have no quantitative meaning (you aren't really able to see the trends that scatterplots are useful for viewing). (Note: Is this right/are there other reasons as well?)
    
    
```{r}
ggplot(data = mpg) +
      geom_point(aes(x = class, y = drv))
```

# 3.3 Aesthetic Mappings

  > aesthetic: a visual property of objects in a plot, which changes levels of such things as size, shape, and color

You may have noticed that there are a few interesting points in the bottom right of the plot of engine displacement against highway miles per gallon. We call them interesting because they are outside of what appears to be a generally linear trend. Why is that occurring?

```{r}
ggplot(data = mpg, aes(displ, hwy)) +
  geom_point()

```

One hypothesis we'll investigate here is this: Perhaps those cars are hybrids. This might make sense since they perform better in highway miles per gallon than other cars with similar engine displacement.

First, we'll see what the ```class``` variable may reveal. For reference, this variable describes what type of car the observation is (i.e. minivan, SUV).
```{r}
ggplot(data = mpg, aes(displ, hwy)) +
  geom_point(aes(color = class))
```

As it turns out, most of these points are two-seater cars (one is a midsize car). The book authors point out here that two-seaters are more efficient for their engine displacement due to smaller bodies -- the distinction wasn't due to the cars being a hybrid after all.

## Exercises
1. Why does this code not produce blue points?

    At a glance, the issue is parentheses placement. Color is being grouped with the aesthetics mapping. Since color doesn't have any connection with the data, it actually shouldn't be used in the aesthetic data mapping.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```


2. Which variables in ```mpg``` categorical? Which are continuous?
    
    The categorical variables in ```mpg``` include manufacturer, model, and transmission (trans), among others. Examples of continuous variables include engine displacement (displ) and city miles per gallon (cty).
```{r}
head(mpg)
```


3. Map a continuous variable to ```color```, ```size```, and ```shape```. How do these aesthetics behave differently for categorical vs continuous variables?
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = -cyl))
  # legend would need to be recalibrated here or need to find color setting to reverse scale
```

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, size = cyl))
  # The scale difference between 4 and 8, or even 4 and 5 doesn't make any sense
```

```{r, eval=FALSE}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, shape = cyl))
  # Doesn't work!
```

4. What happens if you map the same variable to multiple aesthetics?

    It works, but some combinations aren't recommended.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = class, size = class))
```

5. What does the ```stroke``` aesthetic do? What shapes does it work with?

    From the docs:
    
    > Use the stroke aesthetic to modify the width of the border

```{r}
?geom_point
```

6. What happens if you map an aesthetic to something other than a variable name?

    My guess before running: only engine displacement from 1 to 4 will be colored?
    
    It actually just colored by whether or not the condition was met. Handy!
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = displ < 5))
```



# 3.5 Facets

  > facet: subplot that displays one subset of the data

You can facet by a single variable using ```facet_wrap()```, which expects a discrete variable.

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ class, nrow = 3)
```


You can facet by two variables using ```facet_grid()```.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ cyl)
```

Plot with shared axis using dot operator (```.```) in place of row/column:
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(class ~ .)
```

## Exercises
1. What happens if you facet on a continuous variable?

    My guess before running: it shouldn't run, given that facetting is based on classifications.
    
    Turns out - it does run, but it doesn't create a particularly useful plot.

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cty)
```

2. What do the empty cells in the plot facetted by ```drv``` and ```cyl```?

    The empty cells show which intersections of the two variables don't have any observations that meet the two conditions.
    
    How do those empty cells relate to the below plot?
    
    You can see those same gaps. Anywhere that was empty by facet doesn't have a point at the intersection of the two classes.
    
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = drv, y = cyl))
```

3. What plots will the following code make? What does ```.``` do?

    The plots will be facetted in a single column or row, respectively.

4. What are the advantages and disadvantages to using facetting instead of color for class? How might these be affected by using a larger dataset?

    One advantage is being able to have a sense of within-class trends. It is harder to compare the exact values between classes, however. I think these trends would hold even with a larger dataset, but, depending on the image dimensions, you could lose the ability to discern what's going on. (Note: What is the recommended answer here?)

5. What doe the options of ```facet_wrap()``` represent? (```nrow```, ```ncol```, other layout options) Why doesn't ```facet_grid()``` have ```nrow``` and ```ncol``` arguments?

    ```facet_wrap()``` is designed to make efficient use of space with one-dimensional sequences. ```nrow``` and ```ncol``` can be used to control how that two-dimensional adjustment occurs. ```facet_grid()``` is predefined as a matrix (i.e. with two dimensions).
```{r}
?facet_wrap
?facet_grid
```

6. When using ```facet_grid()``` you should usually put the variable with more unique levels in the columns. Why?

    The more unique values, the longer or wider a plot would be when facetted by a given dimension.

# 3.6 Geometric objects

> geom: the visual object used to represent the data

## Exercises
1. What geom would you use to draw... 

    a line chart?
    ```geom_line()```
    
    a boxplot?
    ```geom_boxplot()```
    
    a histogram?
    ```geom_histogram()```
    
    an area chart?
    ```geom_area```

2. What do you think this code will produce? What does it actually produce?
    
    ```
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
      geom_point() + 
      geom_smooth(se = FALSE)```
 
    
    Prediction: a plot of engine displacement and highway miles per gallon that shows both a scatterplot of the data and a smoothed line through the data (without a standard error band), colored by drivetrain type

    Corrections - smoothed lines for each type of drivetrain
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)
```

3. What does ```show.legend = FALSE``` do? What happens if you remove it? Why do you think it was used earlier in the chapter?

    This is the option to hide the legend. If removed, the legend would be displayed. In the triptych of plots earlier in the chapter, the other plots didn't have legends, so this helped the final plot have a similar appearance.
 
4. What does the ```se``` argument to ```geom_smooth()``` do?

    This argument removes the standard error bands around the smoothed line.

5. Will these two graphs look different? Why or why not?

```{r, eval = FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

    Since both are mapping the using the same data with the same x/y mappings, the graphs will be the same.

6. Recreate the R code necessary to generate the following graphs.
```{r, fig.height=2, fig.width=4}
d = ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point()

d +
  geom_smooth(se = FALSE)

d +
  geom_smooth(mapping = aes(group = drv), se = FALSE)

d1 = ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) +
      geom_point()

d1 +
  geom_smooth(se = FALSE)

d1 +
  geom_smooth(se = FALSE, mapping = aes(color = NULL))

d1 +
  geom_smooth(se = FALSE, mapping = aes(linetype = drv, color = NULL))


ggplot(data = mpg, mapping = aes(x = displ, y = hwy, fill = drv)) +
  geom_point(shape = 21, color = 'white', stroke = 2, size = 3)
```

# 3.7 Statistical Transformations
Let's start with a new dataset ```diamonds```. In the graph below, we see that there are many more 'ideal', 'premium', and 'very good' diamonds in this dataset (more high quality than low quality cuts).
```{r}
ggplot(data = diamonds, aes(x = cut)) +
  geom_bar()
```

Bar charts are an example of a chart which maps data beyond the raw values present in the data set (compared to, for example a scatterplot).

> stat: statistical transformation; the algorithm used to calculate new values for your graph

Each geom has a default stat, and each stat has a default geom. This means that you can use them interchangably for simple plots. Three cases where you might need to explicitly state the stat:

1. Overriding the default statistical transformation
        (i.e. manually adding in your own values)
2. Overriding the default mapping from transformed variables to aesthetics
        (i.e. a bar chart of proportions instead of counts)
3. Showing additional detail (within code) about the statistical transformation used

## Exercises

1. What is the default geom used with ```stat_summary()```?
    ```pointrange```

    Show a plot using the default geom instead of ```stat_summary
    ```.

    (Couldn't quite figure this one out. Solved on Stack Overflow [here](https://stackoverflow.com/questions/41850568/r-ggplot2-pointrange-example). The real thing I need to find out is how to dynamically ID min/max by group, rather than by hardcoding as in some other examples.)

````{r, eval=FALSE}
ggplot(data = diamonds, aes(x = cut, y = depth)) +
  #geom_pointrange(mapping = aes(ymin = min(depth), ymax = max(depth)))
  #geom_linerange(mapping = aes(ymin = depth - 1, ymax = depth + 1))
````


2. What does ```geom_col()``` do? How is it different than ```geom_bar()```?

    ```geom_col()``` uses the ```stat_identity()``` whereas ```geom_bar()``` uses ```stat_count``` by default. This means that ```geom_bar()``` will show summarized values, while ```geom_col()``` will show the actual data values.
    

3. Most geoms and stats come in pairs that are almost always used in concert. Read through the documentation and make a list of all the pairs. What do they have in common?

    After surveying most of the geometric objects, you see that most of those with a companion stat are named similarly, using a pattern ```geom_<object>``` : ```stat_<object>```. For example, ```geom_contour()``` : ```stat_contour()```.
    
    
    

- Reference lines: ```geom_abline()``` : ```-``` 

    (also ```geom_hline()```, ```geom_vline()```)
    
- Bar charts: ```geom_bar()``` : ```stat_count()```

- Column charts: ```geom_col()``` : ```-```

- Heatmap of 2d bin counts: ```geom_bin2d()``` : ```stat_bin_2d()```

- Box and whisker plot: ```geom_boxplot()``` : ```stat_boxplot()```

- 2d contours of 3d surface: ```geom_contour()``` : ```stat_contour()```

- Count overlapping points: ```geom_count()``` : ```stat_sum()```
    
    (looks kind of like a bubble chart; example [here](http://r4ds.had.co.nz/visualize.html).)

- Smoothed density estimates: ```geom_density()``` : ```stat_density()```

- Contours of a 2d density estimate: ```geom_density_2d()``` : ```stat_density_2d()```

- Dotplot: ```geom_dotplot()``` : ```-``` 

- Horizontal error bars: ```geom_errorbarh``` : ```-``` 

- Hexagonal heatmap of 2d bin counts: ```geom_hex()``` : ```stat_bin_hex()```

- Histograms and frequency polygons: ```geom_histogram()``` / ```geom_freqpoly()``` : ```stat_bin()```

- Jittered points: ```geom_jitter()``` : ```-``` 

- Vertical intervals (lines, crossbars, and errorbars):
  
    ```geom_crossbar()```
    ```geom_errorbar()```
    ```geom_linerange()```
    ```geom_pointrange()```

- Polygons from a reference map: ```geom_map()``` : ```-``` 

- Connect observations:
  
    ```geom_path()```
    ```geom_line()```
    ```geom_step()```


- Points (scatterplot): ```geom_points()``` : ```-```

- Polygons: ```geom_polygon()``` : ```-```

- Quantile-quantile plots: ```geom_qq()``` : ```stat_qq()```

- Quantile regression: ```geom_quantile()``` : ```stat_quantile()```

- Ribbons and area plots: 

    ```geom_ribbon()```
    ```geom_area()```

- Rug plots: ```geom_rug()``` : ```-```

- Line segments and curves: 

    ```geom_segment()```
    ```geom_curve()```

- Smoothed conditional means: ```geom_smooth()``` : ```stat_smooth()```

- Line segments parameterized by location, direction, and distance: ```geom_spoke()``` : ```-```

- Text: 

    ```geom_text()```
    ```geom_label()```

- Rectangles: 

    ```geom_rect()```
    ```geom_tile()```
    ```geom_raster()```

- Violin plot: ```geom_violin()``` : ```stat_ydensity()```

4. What variables does ```stat_smooth()``` compute? What parameters control its behavior?

    This function computes y as a (some) function of x, and can be conrolled with such items as what type of formula to use for smoothing and what smoothing method to use.

5. In our proportion bar chart, we need to set ```group = 1```. Why? In other words, what is the problem with these two graphs?

    The group setting is used to override default grouping behavior. This allows for the proportion to be calculated across all groups instead of within each group. See a more detailed answer on [Stack Overflow](https://stackoverflow.com/questions/39878813/r-ggplot-geom-bar-meaning-of-aesgroup-1).

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = ..prop..))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = color, y = ..prop..))
```


# 3.8 Position adjustments
Available options:
- ```position_dodge``` (dodge overlapping objects side-to-side)
- ```position_identity``` (don't adjust position)
- ```position_jitter``` (jitter points to avoid overplotting)
- ```position_jitterdodge``` (simultaneously dodge and jitter)
- ```position_nudge``` (nudge points a fixed distance)
- ```position_stack``` | ```position_fill``` (stack overlapping objects on top of each other)

## Exercises
1. What is the problem with this plot? How could you improve it?

    This plot likely suffers from overplotting and could be improved with jitter added in.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point()
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point(position = 'jitter')
```

2. What parameters to ```geom_jitter()``` control the amount of jittering?

    Jittering is controlled by ```width``` and ```height``` in this function.

3. Compare and contrast ```geom_jitter()``` with ```geom_count()```.

    Both functions attempt to shed clarity when points are being overplotted. ```geom_jitter()``` improves the ability to highlight trends, while ```geom_count()``` improves the ability to identify membership at a given intersection point.

4. What's the default position adjustment for ```geom_boxplot()```? Create a visualization of the ```mpg``` dataset that demonstrates it.

```{r}
ggplot(data = mpg, aes(x = drv, y = hwy)) +
  geom_boxplot() # default position is dodge
```


# 3.9 Coordinate systems
Determines how `x` and `y` aesthetics combine to position elements in a plot.

Available options:
- `coord_cartesian` (default)
- `coord_fixed` (Cartesian coordinates with fixed 'aspect ratio')
- `coord_flip` (Cartesian coordinates with x and y flipped)
- `coord_map` and `coord_quickmap` (map projections)
- `coord_polar` (polar coordinates)
- `coord_trans` (transformed Cartesian coordinate system)

## Exercises
1. Turn a stacked bar chart into a  pie chart using ```coord_polar()```.

```{r}
ggplot(data = mpg, aes(x = factor(1), fill = factor(class))) +
  geom_bar(width = 1) +
  coord_polar(theta = 'y')
```

2. What does `labs()` do?

    From the documentation: This function allows you to change the axis, legend, and plot labels, including titles, subtitles, and captions.