---
title: "Chapter 3"
output:
  html_notebook: 
    highlight: tango
    theme: spacelab
    code_folding: show
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, include = TRUE)
```


```{r, include=FALSE}
library(tidyverse)
library(knitr)
```

# 3.1 First Steps
```{r view}
head(mpg) # preview data
```


```{r plot}
ggplot(data = mpg) +
    geom_point(mapping = aes(x = displ, y = hwy))
```


##Exercises
1. What happens with the following code ```ggplot(data = mpg)```?

    Nothing - the ggplot() function just creates the coordinate system for later plots.

2. How many rows are in mpg? How many columns?
```{r}
nrow(mpg)
```
```{r}
ncol(mpg)
```

3. What does the ```drv``` variable describe?
      
      ```drv``` describes whether or not a car has front, rear, or four wheel drive.

```{r}
?mpg
```

4. Make a scatterplot of hwy vs cyl.
```{r}
ggplot(data = mpg) +
  geom_point(aes(x = cyl, y = hwy))
```

5. What happens if you make a scatterplot of ```class``` and ```drv```? Why is the plot not useful?

    The plot isn't useful as a scatterplot, because it is mapping two categorical variables against one another on a plot that is intended to represent quantitative scale -- the differences between steps have no quantitative meaning.
    
    
```{r}
ggplot(data = mpg) +
      geom_point(aes(x = class, y = drv))
```
